---
title: "Exploring Gapminder Dataset (HW2 - STAT 545)"
author: "Phuong (Sam) Can"
date: "24/09/2019"
output:
  pdf_document:
    toc: yes
  html_document:
    keep_md: yes
    toc: yes
  toc_float: yes
---

Let's tackle each question in the assignment:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(gapminder)
str(gapminder)
attach(gapminder)
```

1.1 & 1.2. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# 1.1 & 1.2
gapminder %>%
  filter(country == sample(country, 3) & year > 1969 & year < 1980) %>%
  select(country, gdpPercap)
```

1.3. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# 1.3
gapminder %>% 
  group_by(country) %>% 
  summarise(lifeExp_increase = (diff(lifeExp, lag = 11) > 0)) %>% 
  filter(lifeExp_increase == TRUE)
```

1.4. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# 1.4
gapminder %>% 
  group_by(country) %>% 
  summarise(max_gdpPercap = max(gdpPercap))
```

1.5. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# 1.5
ggplot(gapminder, aes(x=lifeExp, y=gdpPercap)) + geom_point() + scale_x_log10(lifeExp)
```

```{r warning=FALSE, message=FALSE, echo=TRUE}
options("scipen"=100, "digits"=4)
```

2. I selected two variables for exploration: *continent* (categorical) and *pop* (quantitative)
- *continent* has ``r nlevels(continent)`` levels, which are: ``r levels(continent)``.
- *pop* has a mean of ``r mean(pop)``, median of ``r median(pop)``, ranging from ``r min(pop)`` to ``r max(pop)``, and has a standard deviation of ``r sd(pop)``. Here is a histogram to describe the distribution of the world population in the dataset.

```{r warning=FALSE, message=FALSE, echo=TRUE}
hist(pop)
```

Here is a boxplot to describe the relationship between these two variables, where you can see the median, quantiles, and ranges of values for population in each continent.

```{r warning=FALSE, message=FALSE, echo=TRUE}
ggplot(gapminder, aes(x=continent, y=pop)) + geom_boxplot() + scale_y_log10()
```

3. Explore various plot types:
a. A scatterplot with two quantitative variables: *lifeExp* and *gdpPercap.*

```{r warning=FALSE, message=FALSE, echo=TRUE}
ggplot(gapminder, aes(lifeExp, gdpPercap)) + geom_point(alpha = 0.15) + theme_bw() 
```

b. A bar graph with one quantitative and one categorical variable: 

```{r warning=FALSE, message=FALSE, echo=TRUE}
ggplot(gapminder, aes(continent, pop)) + geom_col()
```

Extra 1. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

They did succeed, because they included the dataset (in the first argument) and correct syntax for the second argument, which involves double equal signs "==" and a vector that contains selected countries' names.

Extra 2. We have the output as below:

```{r warning=FALSE, message=FALSE, echo=TRUE}
knitr::kable(head(gapminder, 10))
DT::datatable(gapminder)
```





